Drought indices of water demand and supply, soil moisture, vegetation, and their impact on LULCC in continental Chile

### 2024-01-22 article/paper_drought_Chile2023.tex
actualizar repo

\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Drought indices of water demand and supply, soil moisture, vegetation, and its impact on LULCC in continental Chile},
  pdfauthor={Francisco Zambrano},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Drought indices of water demand and supply, soil moisture,
vegetation, and its impact on LULCC in continental Chile}
\author[1]{Francisco Zambrano%
\corref{cor1}%
\fnref{fn1}}
 \ead{francisco.zambrano@umayor.cl}

@@

\begin{abstract}
Human-induced greenhouse gas emissions have increased the frequency
and/or intensity of weather and climate extremes. Central Chile has been
affected by a persistent drought which is impacting the hydrological
system and vegetation development. The region has been the focus of
research studies due to the diminishing water supply, this persistent
period of water scarcity has been defined as a ``mega drought''.
Nevertheless, our results evidence that the water deficit has expanded
beyond. Our goal is to analyze the impact of drought, measured by
drought indices of water supply/demand and vegetation status, in the
LULCC (land use land cover change) over continental Chile. For the
analysis, continental Chile was divided into five zones according to a
latitudinal gradient: ``Norte Grande'', ``Norte Chico'', ``Zona
Central'', ``Zona Sur'', and ``Zona Austral''. We used monthly climatic
re-analysis variables for precipitation, temperature and soil moisture
for 1981-2023 from ERA5-Land (ERA5L); and MODIS (Moderate-Resolution
Imaging Spectroradiometer) product MCD12Q1 for land cover for 2001-2021,
and the NDVI vegetation index from product MOD13A2 collection 6.1 for
2000-2023, both from collection 6.1. We estimated atmospheric
evaporative demand (AED) by combining the Hargreaves-Samani equation
with the ERA5L temperature. We derived the drought indices SPI
(Standardized Precipitation Index), SPEI (Standardized Precipitation
Evapotranspiration Index), EDDI (Evaporative Demand Drought Index), SSI
(standardized anomaly of cumulative soil moisture), and the zcNDVI
(standardized anomaly of cumulative NDVI). These indices were calculated
for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI (1,
3, and 6 months). We analyzed the temporal correlation of SPI, SPEI,
EDDI, and SSI with zcNDVI to have insights into the impact of water
supply and demand on vegetation. Our results showed that LULCC had an
increasing trend of 412 {[}km2yr−1{]} of forest expansion in the ``Zona
Sur'', together with a decreasing trend of 24 {[}km2yr−1{]} of cropland
contraction in the ``Zona Central'' meanwhile the ``Zona Sur'' showed an
increase of 31 {[}km2yr−1{]}, and a contraction of 80 {[}km2yr−1{]} of
bare soil in the ``Zona Austral''. The EDDI was the less correlated
index for the five macro zones and the five types of land cover, showing
that the temperature in Chile has little impact on vegetation. Higher
r-squared values, between 0.5 and 0.8, were obtained at ``Norte Chico''
and ``Zona Central'' for the land cover types of savanna, shrubland,
grassland, and croplands for the indices SPEI and SSI at time scales of
12 and 24 months. The forest type reaches a r-squared of
\textasciitilde0.5 for SSI of 12 months. The results indicate that the
``Norte Chico'' and ``Zona Central'' are the most sensitive regions to
water supply deficits longer than a year, potentially explained by a low
capacity of water storage in those zones that should be further
investigated.
\end{abstract}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, frame hidden, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi

@@

The sixth assessment report (AR6) of the IPCC \citep{IPCC2023} indicates
that human-induced greenhouse gas emissions have increased the frequency
and/or intensity of some weather and climate extremes, and the evidence
has been strengthened since AR5 \citep{IPCC2013}. There is high
confidence that increasing global warming can expand the land area
affected by increasing drought frequency and severity
\citep{IPCCCH112021}. Furthermore, drought increases tree mortality and
triggers changes in land cover and, consequently, land use, thus
impacting ecosystems \citep{Crausbay2017}. Nevertheless, there is a lack
of understanding of how the alteration in water supply and demand is
affecting land cover transformations.

Precipitation is the primary driver of drought and is intensified by
temperature \citep{Luo2017}. Drought impacts soil moisture, hydrological
regimes, and vegetation productivity. Initially, drought was commonly
classified as meteorological, hydrological, and agricultural
\citep{Wilhite1985}. Lately, \citet{Loon2016} and
\citet{AghaKouchak2021} have given an updated definition of drought for
the Anthropocene, suggesting that it should be considered the feedback
of humans' decisions and activities that drives the anthropogenic
drought. Even though it has been argued that those definitions do not
fully address the ecological dimensions of drought. \citet{Crausbay2017}
proposed the ecological drought definition as \emph{``an episodic
deficit in water availability that drives ecosystems beyond thresholds
of vulnerability, impacts ecosystem services, and triggers feedback in
natural and/or human systems''}. Moreover, many ecological studies have
misinterpreted how to characterize drought, for example, sometimes
considering ``dry'' conditions as ``drought'' \citep{Slette2019}. On the
other hand, the AR6 \citep{IPCC2023} states that even if global warming
is stabilized at 1.5°--2°C, many parts of the world will be impacted by
more severe agricultural and ecological droughts. Then, there is a
challenge in conducting drought research, especially to evaluate its
impact on ecosystems.

@@

For the spatiotemporal assessment of drought impact (i.e., by water
supply and demand) on land cover changes, we need climatic realiable
variables such as precipitation, temperature, soil moisture, land cover,
and vegetation status. For developing countries like Chile, the weather
networks present several disadvantages, such as gaps, a short history,
and low-quality data. Reanalysis data, as the ERA5-Land (ERA5L)
\citep{MunozSabater2021} provides hourly climatic information
(precipitation, temperature, and soil moisture) without gaps since 1950
with global extension. ERA5L has already been used for drought
assessment using the Standardized Precipitation-Evapotranspiration Index
(SPEI) \citep{Nouri2023} and for flash drought \citep{Wang2023} by
analyzing soil mositure and evapotranspiration. On the other hand,
satellite remote sensing \citep{West2019, AghaKouchak2015} is the
primary method to evaluate how drought impacts vegetation dynamics.
Vegetation drought indices (VDI) are commonly used as proxies of
productivity \citep{Paruelo2016, Schucknecht2017}, which can be derived
from the MODIS (Moderate-Resolution Imaging Spectroradiometer). Besides,
land use and land cover (LULC) change can be driven by drought
\citep{Tran2019, Akinyemi2021}. To analyze these changes, multiple LULC
products exist \citep{Grekousis2015}, one of those that provides time
series since 2001 is the MCD12Q1 \citep{Friedl2019} from MODIS. The
variation in water supply and demand is finally reflected in the total
water storage (TWS). The TWS can be retrieved by the Gravity Recovery
and Climate Experiment (GRACE), which allows analyzing water
availability changes at coarse resolution \citep{Ahmed2014, Ma2017}. We
can use climatic reanalysis (ERA5L) and vegetation data (MODIS) to
derive drought indices of supply (i.e., precipitation) and demand (i.e.,
temperature) and thus evaluate the impact of drought on LULC changes.
Further, the TWS can be assessed with regard to the changes in water
supply and demand to gain insight into the impact on water storage.

@@

In this research, we present the raster dataset DDS4Chl, which provides
climate variables and drought indices of water demand and supply and
vegetation productivity for continental Chile since 1981 at a monthly
frequency. Those were gathered from the earth observation products ERA5L
and MODIS. Then, we used DDS4Chl to analyze the impact of drought on
different types of land cover classes in continental Chile. The specific
objectives of the study are: i) to analyze the trend on multi-scalar
drought indices for water demand and supply, soil moisture, and
vegetation productivity for 1981--2023/2001--2023; ii) to evaluate LULC
change for 2001--2021 and its relation to drought indices; iii) to
analyze the relationship of a proxy of vegetation productivity (zcNDVI)
with drought indices of water demand and supply and soil moisture; and
iv) to assess if the observed changes in the drought indices are linked
to TWS.

@@

Continetal Chile has a diverse climate condition from north to south and
east to west \citep{Aceituno2021} (Figure~\ref{fig-studyArea}), which
determines its great ecosystem diversity (Figure~\ref{fig-LCprop}). The
Andes Mountains are a main factor in latitudinal variation
\citep{Garreaud2009}. To describe the climate and ecosystem of Chile, we
use the Koppen-Geiger release by \citet{Beck2023} and the landcover type
persistance of 80\% of times for 2001--2022, from the IGBP
classification scheme \citep{Friedl2019} (see
Section~\ref{sec-methods_lulc}). ``Norte Grande'' and ``Norte Chico''
predominate in an arid desert climate with hot (Bwh) and cold (Bwk)
temperatures. At the south of ``Norte Chico,'' the climate changes to an
arid steppe with cold temperatures (Bsk). Mainly, the land is barren,
with a minor surface of vegetation types such as shrubland and
grassland. In the zones ``Centro'' and the north half of ``Sur,'' the
main climate is Mediterranean, with warmer to hot summers (Csa and Csb).
In ``Centro,'' there is a major proportion (\textasciitilde50\%) of
Chilean matorral (shrubland and savanna, \citep{Fuentes2021}), followed
by grassland (\textasciitilde16\%), forest (8\%), and croplands (5\%).
The south part of ``Sur'' and the north part of ``Austral'' are
dominated by an Oceanic climate (Cfb). Those zones are high in forest
and grassland. The southern part of the country has a tundra climate,
and in Patagonia, it is a cold semi-arid area with an extended surface
of grassland, forest, and, to a lesser extent, savanna.

@@

\begin{table}[!ht]
\caption{Description of the earth observation data used }
\label{tab-desEOD}
\small
\centering
\begin{tabular}{p{0.13\textwidth}cp{0.3\textwidth}p{0.095\textwidth}ccc}
\hline
\multirow{1}{*}{\centering Product} & Sub-product & Variable & Spatial Resolution  & Period & Units & Short Name \\ 
\hline
\multirow{4}{*}{ERA5L} & ~ & Precipitation & \multirow{4}{*}{~0.1°} & \multirow{4}{*}{1981-2023} & mm & P \\ 
         &  & Maximum temperature & ~ & & $°C$ & $T_{max}$ \\ 
         &  & Minimum temperature & ~ & & $°C$ & $T_{min}$ \\ 
         &  & Volumetric Soil Water Content at 1m & ~ & & $m3/m3$ & SM \\ 
ERA5L* & & Atmospheric Evaporative Demand & 0.1° & 1981-2023 & mm & AED \\
        \multirow{2}{*}{MODIS} & MOD13A3.061 & Normalized Difference Vegetation Index & \multirow{2}{*}{~1 km} & 2000-2023 & ~ & NDVI \\ 
         & MCD12Q1.061 & Landcover IGBP scheme & & 2001-2022 & ~ & Landcover \\ 
\hline
\end{tabular}
{\raggedright *Derived from ERA5L with Eq. \ref{eq-AED}. \par}
\end{table}

@@

\hypertarget{landcover-macroclasess-and-validation}{%
\subsubsection{Landcover macroclasess and
validation}\label{landcover-macroclasess-and-validation}}

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
6.1 from MODIS. This product has a yearly frequency from 2001 to 2022.
The IGBP defines 17 classes; from these, we regrouped into ten
macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9
to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands,
13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
Thus, we have a landcover raster time-series with the ten classes for
2001 and 2023.

To validate the landcover obtained, we compare the macroclasses with the
ones of a more detailed landcover map made by \citet{Zhao2016} for Chile
with samples acquired in the years 2013--2014 (LCChile). The later has a
spatial resolution of 30 m and three levels of defined classes; from
those, we used level 1, which fits with the macroclasses landcover. We
chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from landcover
macrolcasses to validate with LCChile.

@@

\begin{enumerate}
\def\labelenumi{\roman{enumi})}
\tightlist
\item
  resampled LCChile to the spatial resolution (500m) of the landcover
  macroclasses using the nearest neighbor method,
\item
  took a random sample of 1000 points within continental Chile and
  extracted the classes that fell within each point for LCChile,
  IGBP2013, and IGBP2014; we considered the point extracted from LCChile
  as the truth and the values as the other two years as prediction
\item
  calculate a confusion matrix with the classes extracted in the 1000
  poitns for LCChile, IGBP2013, and IGBP2014. Calculate the performance
  metrics of accuracy and F1.
\end{enumerate}

@@

\hypertarget{landcover-persistence-mask-2001-2022}{%
\subsubsection{Landcover persistence mask
2001-2022}\label{landcover-persistence-mask-2001-2022}}

The time series of NDVI is affected by climatic conditions, vegetation
development, seasonality, and changes in vegetation type. In this study,
we want to analyze the variation in vegetation productivity in different
landcover types and how it is affected by water demand, water supply,
and soil moisture. In order to avoid changes due to a change in the
landcover type, that will wrongly impact NDVI. We will develop a
persistence mask for landcover for 2001--2023. Thereby, we reduce an
important source of variation on a regional scale.

Thus, we calculated a raster mask for IGBP MODIS considering the
macroclasses that remain without change for more than 80\% of the years
(2001--2022) per pixel, which allows us to identify the areas with no
landcover change for the macroclasses.

\hypertarget{landcover-trend-and-drought-indices}{%
\subsubsection{Landcover trend and drought
indices}\label{landcover-trend-and-drought-indices}}

We calculated the surface occupied per landcover class into the five
macrozones (``Norte Grande'' to ``Austral'') per year for 2001--2023.
After that, we calculated the trend's change in surface; we used the
Sen' slope \citep{Sen1968} based on Mann-Kendall \citep{Kendall1975}.
This way, we obtain a matrix of trends of 5 x 5 (macrozones x
landcover). The aim is to later explore if the trend in landcover
classes is associated with a trend in the drought indices. For this, we
will use the techniques of regresion and regularization of Lasso
\citep{Tibshirani2010} and Ridge \citep{Hoerl1970}. Also, we will test
random forests for this purpose \citep{Ho1995}. We will choose the trend
of landcover surface per macroclass and macrozone as the response
variable and the trend of the drought indices (SPI, SPEI, EDDI, and SSI
for time scales 1, 3, 6, 12, 24, and 36 months) as the predictor
variables. With this analysis, we expect to gather insights regarding
whether there is a pattern of climatic influence along Chile or if what
is happening in Central Chile has to do with more localized climatic
conditions.

@@

To estimate if there are significant positive or negative trends for the
drought indices, we used the non-parametric test of Mann-Kendall
\citep{Kendall1975}. To determine the magnitude of the trend, we used
Sen's slope \citep{Sen1968}. Some of the advantages of applying this
methodology are that the Sen's slope is not affected by outliers as
regular regression does, and it is a non-paramteric method that is not
affected by the distribution of the data. We applied both to the six
time scales from 1981 to 2023 (monthly frequency) and the indices SPI,
EDDI, SPEI, and SSI. In the case of zcNDVI (six months) was for 2000 to
2023. Thus, we have 31 trends. Also, we extracted the trend aggregated
by macrozone and landcover class, obtaining a table of 31x5x5 (drought
indices trends x macrozone x landcover class). We will use this data in
Section~\ref{sec-methods_lulc} to analyze if there is a strong
relationship between the trends of drought indices and land cover
surface within continental Chile.

>>>>

\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Drought indices of water demand and supply, soil moisture, vegetation, and their impact on LULCC in continental Chile},
  pdfauthor={Francisco Zambrano},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Drought indices of water demand and supply, soil moisture,
vegetation, and their impact on LULCC in continental Chile}
\author[1]{Francisco Zambrano%
\corref{cor1}%
\fnref{fn1}}
 \ead{francisco.zambrano@umayor.cl}

@@

\begin{abstract}
Central Chile has been the focus of research studies due to the
persistent decrease in water supply, which is impacting the hydrological
system and vegetation development. This persistent period of water
scarcity has been defined as a ``Mega Drought''. Our objective is to
examine the effects of drought on LULCC (land use land cover change)
over continental Chile using drought indices of water supply and demand,
soil moisture, and vegetation productivity. For the analysis,
continental Chile was divided into five zones according to a latitudinal
gradient: ``Norte Grande,'' ``Norte Chico,'' ``Centro,'' ``Sur,'' and
``Austral.'' The monthly ERA5-Land (ERA5L) variables for precipitation,
temperature, and soil moisture were used. From 2001 to 2022, we used the
land cover MODIS product MCD12Q1, and from 2000 to 2023, we used the
NDVI (Normalized Difference Vegetation Index) product MOD13A3 collection
6.1. We estimated atmospheric evaporative demand (AED) using the
Hargreaves-Samani equation with the ERA5L temperature. We used the
Standardized Precipitation Index (SPI), the Standardized Precipitation
Evapotranspiration Index (SPEI), the Evaporative Demand Drought Index
(EDDI), the Standardized Soil Moisture Index (SSI), and the Standardized
anomaly of cumulative NDVI (zcNDVI) as drought indicators. These indices
were calculated for time scales of 1, 3, 6, 12, 24, and 36 months,
except for zcNDVI, which was for 6 months. We analyze the trend for
LULCC, vegetation productivity, and drought indices. Also, we analyzed
the temporal correlation of SPI, SPEI, EDDI, and SSI with zcNDVI to gain
insights into the impact of water supply and demand on vegetation
productivity. Our results showed that LULCC were highest in ``Centro,''
``Sur,'' and ``Austral,'' with 36\%, 31\%, and 34\%, respectively. The
EDDI shows that water demand has increased for all zones, with a major
increase in ``Norte Grande.'' The drought indices of water supply and
soil moisture evidence a decreasing trend, which decreases at longer
time scales, from ``Norte Grande'' to ``Sur.'' ``Austral'' is the only
zone that shows an increase in supply. Vegetation productivity measures
by zcNDVI present a negative trend in ``Norte Chico'' and ``Centro.''
Showing to be the zones most impacted by climatic conditions, the years
2020 and 2022 suffered the most extreme drought. On the other hand,
forests seem to be the most resistant to drought. The types that show to
be most affected by variation in climate conditions are shrublands,
savannas, and croplands. The drought indices that have the capability of
explaining to a major degree the variance in vegetation productivity are
SSI-12, followed by SPEI-24 and SPEI-12 in ``Norte Chico'' and
``Centro.'' The results indicate that ``Norte Chico'' and ``Zona
Central'' are the most sensitive regions to water supply deficits longer
than a year, potentially explained by a low capacity of water storage in
those zones that should be further investigated.
\end{abstract}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, sharp corners, frame hidden, interior hidden, enhanced]}{\end{tcolorbox}}\fi

@@

The sixth assessment report (AR6) of the IPCC \citep{IPCC2023} indicates
that human-induced greenhouse gas emissions have increased the frequency
and/or intensity of some weather and climate extremes, and the evidence
has been strengthened since AR5 \citep{IPCC2013}. There is a high degree
of confidence that rising temperatures will increase the land area where
droughts will occur more frequently and with greater severity
\citep{IPCCCH112021}. Furthermore, drought increases tree mortality and
triggers changes in land cover and, consequently, land use, thus
impacting ecosystems \citep{Crausbay2017}. Nevertheless, there is a lack
of understanding of how the alteration in water supply and demand is
affecting land cover transformations.

The primary cause of drought is precipitation, and temperature makes it
worse \citep{Luo2017}. Drought impacts soil moisture, hydrological
regimes, and vegetation productivity. Initially, drought was commonly
classified as meteorological, hydrological, and agricultural
\citep{Wilhite1985}. Lately, \citet{Loon2016} and
\citet{AghaKouchak2021} have given an updated definition of drought for
the Anthropocene, suggesting that it should be considered the feedback
of humans' decisions and activities that drives the anthropogenic
drought. Even though it has been argued that those definitions do not
fully address the ecological dimensions of drought, \citet{Crausbay2017}
proposed the ecological drought definition as ``an episodic deficit in
water availability that drives ecosystems beyond thresholds of
vulnerability, impacts ecosystem services, and triggers feedback in
natural and/or human systems.''. Moreover, many ecological studies have
misinterpreted how to characterize drought, for example, sometimes
considering ``dry'' conditions as ``drought'' \citep{Slette2019}. On the
other hand, the AR6 \citep{IPCC2023} predicts that many regions of the
world will experience more severe agricultural and ecological droughts
even if global warming stabilizes at 1.5°--2°C. Then, there is a
challenge in conducting drought research, especially to evaluate its
impact on ecosystems.

@@

For the spatiotemporal assessment of drought impact (i.e., by water
supply and demand) on land cover changes, we need climatic realiable
variables such as precipitation, temperature, soil moisture, land cover,
and vegetation status. For developing countries like Chile, the weather
networks present several disadvantages, such as gaps, a short history,
and low-quality data. Reanalysis data, as the ERA5-Land (ERA5L)
\citep{MunozSabater2021} provides hourly climatic information
(precipitation, temperature, and soil moisture) without gaps since 1950
with global extension. ERA5L has already been used for drought
assessment using the Standardized Precipitation-Evapotranspiration Index
(SPEI) \citep{Nouri2023} and for flash drought \citep{Wang2023} by
analyzing soil moisture and evapotranspiration. On the other hand,
satellite remote sensing \citep{West2019, AghaKouchak2015} is the
primary method to evaluate how drought impacts vegetation dynamics. The
Moderate-Resolution Imaging Spectroradiometer (MODIS) can be used to get
vegetation drought indices (VDI), which are often used as proxies for
productivity \citep{Paruelo2016, Schucknecht2017}. Besides, land use and
land cover (LULC) change can be driven by drought
\citep{Tran2019, Akinyemi2021}. To analyze these changes, multiple LULC
products exist \citep{Grekousis2015}. One of those that provides time
series since 2001 is the MCD12Q1 \citep{Friedl2019} from MODIS. The
variation in water supply and demand is finally reflected in the total
water storage (TWS). The Gravity Recovery and Climate Experiment
(GRACE), which allows analyzing changes in water availability at coarse
resolution, can retrieve the TWS \citep{Ahmed2014, Ma2017}. We can find
drought indices of supply (i.e., precipitation) and demand (i.e.,
temperature) using climatic reanalysis (ERA5L) and vegetation data
(MODIS). This lets us figure out how drought changes LULC. Further, the
TWS can be assessed with regard to the changes in water supply and
demand to gain insight into the impact on water storage.

@@

In this research, we aim to analyze the impact of drought on different
types of ecosystems (land cover classes) in continental Chile. Our
specific goals are: i) to analyze the trend on multi-scalar drought
indices for water demand and supply, soil moisture, and vegetation
productivity for 1981--2023/2001--2023; ii) to assess the LULC change
for 2001--2021 and how it relates to drought indices; iii) to evaluate
the relationship between zcNDVI, a measure of vegetation productivity,
and drought indices for water demand and supply and soil moisture; and
iv) to assess if the observed changes in the drought indices are linked
to TWS.

@@

Continetal Chile has a diverse climate condition from north to south and
east to west \citep{Aceituno2021} (Figure~\ref{fig-studyArea}), which
determines its great ecosystem diversity (Figure~\ref{fig-LCprop}). The
Andes Mountains are a main factor in latitudinal variation
\citep{Garreaud2009}. To describe the climate and ecosystem of Chile, we
use the Koppen-Geiger release by \citet{Beck2023} and the land cover
type persistance of 80\% of times for 2001--2022, from the IGBP
classification scheme \citep{Friedl2019} (see
Section~\ref{sec-methods_lulc}). ``Norte Grande'' and ``Norte Chico''
predominate in an arid desert climate with hot (Bwh) and cold (Bwk)
temperatures. At the south of ``Norte Chico,'' the climate changes to an
arid steppe with cold temperatures (Bsk). Mainly, the land is barren,
with a minor surface of vegetation types such as shrubland and
grassland. In the zones ``Centro'' and the north half of ``Sur,'' the
main climate is Mediterranean, with warmer to hot summers (Csa and Csb).
There is a significant amount (50\%) of Chilean matorral (shrubland and
savanna, \citep{Fuentes2021}), then grassland (16\%), forest (8\%), and
croplands (5\%), in ``Centro.'' The south part of ``Sur'' and the north
part of ``Austral'' are dominated by an oceanic climate (Cfb). Those
zones are high in forest and grassland. The southern part of the country
has a tundra climate, and in Patagonia, it is a cold semi-arid area with
an extended surface of grassland, forest, and, to a lesser extent,
savanna.

@@

\begin{table}[!ht]
\caption{Description of the earth observation data used }
\label{tab-desEOD}
\small
\centering
\begin{tabular}{p{0.13\textwidth}cp{0.3\textwidth}p{0.095\textwidth}ccc}
\hline
\multirow{1}{*}{\centering Product} & Sub-product & Variable & Spatial Resolution  & Period & Units & Short Name \\ 
\hline
\multirow{4}{*}{ERA5L} & ~ & Precipitation & \multirow{4}{*}{~0.1°} & \multirow{4}{*}{1981-2023} & mm & P \\ 
         &  & Maximum temperature & ~ & & $°C$ & $T_{max}$ \\ 
         &  & Minimum temperature & ~ & & $°C$ & $T_{min}$ \\ 
         &  & Volumetric Soil Water Content at 1m & ~ & & $m3/m3$ & SM \\ 
ERA5L* & & Atmospheric Evaporative Demand & 0.1° & 1981-2023 & mm & AED \\
        \multirow{2}{*}{MODIS} & MOD13A3.061 & Normalized Difference Vegetation Index & \multirow{2}{*}{~1 km} & 2000-2023 & ~ & NDVI \\ 
         & MCD12Q1.061 & land cover IGBP scheme & & 2001-2022 & ~ & land cover \\ 
\hline
\end{tabular}
{\raggedright *Derived from ERA5L with Eq. \ref{eq-AED}. \par}
\end{table}

@@

\hypertarget{land-cover-macroclasess-and-validation}{%
\subsubsection{land cover macroclasess and
validation}\label{land-cover-macroclasess-and-validation}}

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
6.1 from MODIS. This product has a yearly frequency from 2001 to 2022.
The IGBP defines 17 classes; from these, we regrouped into ten
macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9
to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands,
13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
Thus, we have a land cover raster time-series with the ten classes for
2001 and 2023.

To validate the land cover obtained, we compare the macroclasses with
the ones of a more detailed land cover map made by \citet{Zhao2016} for
Chile with samples acquired in the years 2013--2014 (LCChile). The later
has a spatial resolution of 30 m and three levels of defined classes;
from those, we used level 1, which fits with the macroclasses land
cover. We chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from land
cover macrolcasses to validate with LCChile.

@@

\begin{enumerate}
\def\labelenumi{\roman{enumi})}
\tightlist
\item
  resampled LCChile to the spatial resolution (500m) of the land cover
  macroclasses using the nearest neighbor method,
\item
  took a random sample of 1000 points within continental Chile and
  extracted the classes that fell within each point for LCChile,
  IGBP2013, and IGBP2014; we considered the point extracted from LCChile
  as the truth and the values as the other two years as prediction
\item
  calculate a confusion matrix with the classes extracted in the 1000
  poitns for LCChile, IGBP2013, and IGBP2014. Calculate the performance
  metrics of accuracy and F1.
\end{enumerate}

@@

\hypertarget{land-cover-persistence-mask-2001-2022}{%
\subsubsection{land cover persistence mask
2001-2022}\label{land-cover-persistence-mask-2001-2022}}

The time series of NDVI is affected by climatic conditions, vegetation
development, seasonality, and changes in vegetation type. In this study,
we want to analyze the variation in vegetation productivity in different
land cover types and how it is affected by water demand, water supply,
and soil moisture. In order to avoid changes due to a change in the land
cover type, that will wrongly impact NDVI. We will develop a persistence
mask for land cover for 2001--2023. Thereby, we reduce an important
source of variation on a regional scale.

Thus, we calculated a raster mask for IGBP MODIS considering the
macroclasses that remain without change for more than 80\% of the years
(2001--2022) per pixel, which allows us to identify the areas with no
land cover change for the macroclasses.

\hypertarget{land-cover-trend-and-drought-indices}{%
\subsubsection{land cover trend and drought
indices}\label{land-cover-trend-and-drought-indices}}

We calculated the surface occupied per land cover class into the five
macrozones (``Norte Grande'' to ``Austral'') per year for 2001--2023.
After that, we calculated the trend's change in surface; we used the
Sen' slope \citep{Sen1968} based on Mann-Kendall \citep{Kendall1975}.
This way, we obtain a matrix of trends of 5 x 5 (macrozones x land
cover). The aim is to later explore if the trend in land cover classes
is associated with a trend in the drought indices. For this, we will use
the techniques of regresion and regularization of Lasso
\citep{Tibshirani2010} and Ridge \citep{Hoerl1970}. Also, we will test
random forests for this purpose \citep{Ho1995}. We will choose the trend
of land cover surface per macroclass and macrozone as the response
variable and the trend of the drought indices (SPI, SPEI, EDDI, and SSI
for time scales 1, 3, 6, 12, 24, and 36 months) as the predictor
variables. With this analysis, we expect to gather insights regarding
whether there is a pattern of climatic influence along Chile or if what
is happening in Central Chile has to do with more localized climatic
conditions.

@@

To estimate if there are significant positive or negative trends for the
drought indices, we used the non-parametric test of Mann-Kendall
\citep{Kendall1975}. To determine the magnitude of the trend, we used
Sen's slope \citep{Sen1968}. Some of the advantages of applying this
methodology are that the Sen's slope is not affected by outliers as
regular regression does, and it is a non-paramteric method that is not
affected by the distribution of the data. We applied both to the six
time scales from 1981 to 2023 (monthly frequency) and the indices SPI,
EDDI, SPEI, and SSI. In the case of zcNDVI (six months) was for 2000 to
2023. Thus, we have 31 trends. Also, we extracted the trend aggregated
by macrozone and land cover class, obtaining a table of 31x5x5 (drought
indices trends x macrozone x land cover class). We will use this data in
Section~\ref{sec-methods_lulc} to analyze if there is a strong
relationship between the trends of drought indices and land cover
surface within continental Chile.
